SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2019-01-15 scipy-1.0/paper.tex
Paul Dubois CiSE edit cleanup

As SciPy, the algorithms library of the ecosystem, began attracting the attention
of large research organizations,
the next generation of graduate students and postdocs were already exploring
other aspects of the computational environment.
In 2000, Prabhu Ramachandran, a PhD student at the Indian Institute of
Technology, started work on a 3D visualization application, based on
Kitware's C++ Visualization Toolkit\cite{schroeder:2006:VTK}, called
Mayavi\footnote{https://web.archive.org/web/20030731122452/http://pythonology.org/success\&story=mayavi}.
In 2001, Fernando Pérez, a graduate student at the University of
Colorado, Boulder, created the IPython interactive shell (that would
eventually blossom into Project Jupyter\cite{Kluyver:2016aa}), built into
IPP, an earlier attempt by Janko Hauser, and including code from
Nathan Gray's LazyPython.
% IPython hosted on SciPy:
% https://mail.python.org/pipermail/scipy-user/2003-March/001455.html
John Hunter, a postdoc
%John got his PhD in 2001, then a post-doc with John Milton (dept of
%Neurology, UChicago?) a couple of years; then research professor
%under Kurt Hecox, whereafter at Tradelink from 2006
at the University
of Chicago, liked the plotting functionality
available in MATLAB, but had problems accessing their laboratory's
license, which was governed by a hardware dongle.  In response, he
wrote a plotting library from scratch, and Matplotlib 0.1 was released
April 2003\footnote{https://mail.python.org/pipermail/python-list/2003-April/193167.html}.

@@

%In April 2006, Paul
%Dubois\footnote{https://mail.python.org/pipermail/numpy-discussion/2006-April/007487.html} wrote
%to the NumPy mailing list:
%\begin{quote}
%IEEE's magazine, Computing in Science and Engineering (CiSE), has asked me
%to put together a theme issue on the use of Python in Science and
%Engineering. I will write an overview to be accompanied by 3-5 articles [. . .]
%that show a diverse set of applications or
%tools, to give our readers a sense of whether or not Pyth%on might be useful
%in their own work.
%\end{quote}
In a May/June 2007 special issue of IEEE Computing in Science and
Engineering, Paul Dubois wrote\cite{dubois2007guest}:
\begin{quote}
LLNL now has many Python-based efforts built from scratch or wrapped around
legacy codes [. . .] hundreds of thousands
of lines of C++, Python, and Fortran 95, all working together just as we hoped,
doing compute-intensive calculations on massively parallel computers.
\end{quote}
While it is now more commonly accepted, Paul took the time to explain to
the 2007 reader that ``interpreted doesn't mean slow or only interactive.''
He also shares his own interest in Python for ``computational steering''
where ``Python serves as the input language to a scientific application, and the actual
computations are performed both in Python itself and in compiled extensions.''
%https://www.computer.org/csdl/mags/cs/2007/03/c3007.html
Using Python for computational steering was one of the earliest uses of Python
in large scientific processing pipelines.
In addition to Paul's overview there were also articles introducing
Python and SciPy \cite{oliphant2007python}, IPython \cite{perez2007ipython},
and  Matplotlib \cite{hunter2007matplotlib}.
In addition to these longer introductory articles, there was a diverse set of
research applications including
systems biology\cite{myers2007python},
astronomy\cite{greenfield2007reaching},
robotics\cite{krauss2007python},
nanophotonics\cite{bienstman2007python},
partial differential equations\cite{mardal2007using},
neuroimaging\cite{millman2007analysis},
geographic information systems\cite{shi2007python}, and
education\cite{backer2007computational, myers2007pythona}.

@@

In the early workshops, recurrent topics indicate the state of development, with
emphasis being placed on the underlying array package, plotting,
parallel processing, acceleration / wrapping, and user interfaces.  By
2004, a significant shift occurred towards application of SciPy to
scientific problems.  The event also started to draw in more keynote
speakers from outside the community, such as Guido van Rossum (creator
of Python, 2006), Ivan Krstić (One Laptop per Child, 2007), Alex
Martelli (Google, 2008), and Peter Norvig (Google Research, 2009).
The SciPy conference went from being a small gathering of
core developers, to a multi-location event, with increased funding, a
published proceedings, and scholarships for attending students (some
of whom are authors on this paper).  In 2010, the conference grew even
further, with an additional event added in India, and the US
conference—now with parallel tracks—moving to Austin, Texas.
While the US SciPy conference kept growing (by ... it had multiple
tracks, student sponsorship, proceedings, and funding), satellite
events were being organized by volunteers elsewhere, such as EuroSciPy
(2008–) and SciPy India (2009–).
Special sessions and minisymposia dedicated to scientific Python began
appearing at many events.
For example, a three-part minisymposium organized for CSE 2009 was
featured in SIAM News\cite{siamcse09}.

When Paul edited the CiSE May/June 2007 special issue, Python already had a
strong enough presence in science and engineering so that editors solicited
the special issue, but it was still niche enough that the average reader
would need additional information in order to decide whether it would
``be useful in their own work.''
The follow-up CiSE March/April 2011 Python for Scientists and Engineers
special issue\cite{millman2011python}, focused more on the core parts
of the scientific Python ecosystem\cite{perez2011python} including
NumPy\cite{vanderwalt2011numpy}, Cython\cite{behnel2011cython},
and Mayavi\cite{ramachandran2011mayavi}.

Python has become so pervasive, there are now domain-specific special issues.
For example, Frontiers in Neuroinformatics has a collection of 25 articles---covering
topics including modeling and simulation, data collection, electrophysiology, visualization,
as well as stimulus generation and presentation---called Python in
neuroscience\footnote{\url{https://www.frontiersin.org/research-topics/8/python-in-neuroscience}}.

In 2012, Perry Greenfield, John Hunter, Jarrod Millman, Travis Oliphant,
and Fernando Pérez founded NumFOCUS\footnote{https://numfocus.org},
 a 501(c)3 public charity whose mission is to
``promote sustainable high-level programming languages, open code development,
and reproducible scientific research.''
While NumFOCUS is language agnostic, many of the early sponsored projects
came from the scientific Python stack.
Today it has over 25 sponsored projects including NumPy, SciPy, IPython, and
Matplotlib.
Among its other projects, NumFOCUS organizes a global network of 
community driven educational program called PyData.

>>>>

As SciPy, the algorithms library of the ecosystem, began attracting the attention
of large research organizations,
the next generation of graduate students and postdocs were already exploring
other aspects of the computational environment.
In 2000, Prabhu Ramachandran, a PhD student at the Indian Institute of
Technology, started work on a 3D visualization application, based on
Kitware's C++ Visualization Toolkit\cite{schroeder:2006:VTK}, called
Mayavi\cite{mayavi-intro}.
In 2001, Fernando Pérez, a graduate student at the University of
Colorado, Boulder, created the IPython interactive shell (that would
eventually blossom into Project Jupyter\cite{Kluyver:2016aa}), built into
IPP, an earlier attempt by Janko Hauser, and including code from
Nathan Gray's LazyPython.
% IPython hosted on SciPy:
% https://mail.python.org/pipermail/scipy-user/2003-March/001455.html
John Hunter, a postdoc
%John got his PhD in 2001, then a post-doc with John Milton (dept of
%Neurology, UChicago?) a couple of years; then research professor
%under Kurt Hecox, whereafter at Tradelink from 2006
at the University
of Chicago, liked the plotting functionality
available in MATLAB, but had problems accessing their laboratory's
license, which was governed by a hardware dongle.  In response, he
wrote a plotting library from scratch, and Matplotlib 0.1 was released
April 2003\cite{matplotlib-rel}.

@@

%In April 2006, Paul
%Dubois\footnote{https://mail.python.org/pipermail/numpy-discussion/2006-April/007487.html} wrote
%to the NumPy mailing list:
%\begin{quote}
%IEEE's magazine, Computing in Science and Engineering (CiSE), has asked me
%to put together a theme issue on the use of Python in Science and
%Engineering. I will write an overview to be accompanied by 3-5 articles [. . .]
%that show a diverse set of applications or
%tools, to give our readers a sense of whether or not Pyth%on might be useful
%in their own work.
%\end{quote}
In a May/June 2007 special issue of IEEE Computing in Science and
Engineering, Paul Dubois wrote\cite{dubois2007guest}:
\begin{quote}
LLNL now has many Python-based efforts built from scratch or wrapped around
legacy codes [. . .] hundreds of thousands
of lines of C++, Python, and Fortran 95, all working together just as we hoped,
doing compute-intensive calculations on massively parallel computers.
\end{quote}
While it is now more commonly accepted, Paul Dubois took the time to explain to
the 2007 reader that ``interpreted doesn't mean slow or only interactive.''
He also shares his own interest in Python for ``computational steering''
where ``Python serves as the input language to a scientific application, and the actual
computations are performed both in Python itself and in compiled extensions.''
%https://www.computer.org/csdl/mags/cs/2007/03/c3007.html
Using Python for computational steering was one of the earliest uses of Python
in large scientific processing pipelines.
In addition to Paul Dubois' overview there were also articles introducing
Python and SciPy \cite{oliphant2007python}, IPython \cite{perez2007ipython},
and  Matplotlib \cite{hunter2007matplotlib}.
There was also a diverse set of
research applications including
systems biology\cite{myers2007python},
astronomy\cite{greenfield2007reaching},
robotics\cite{krauss2007python},
nanophotonics\cite{bienstman2007python},
partial differential equations\cite{mardal2007using},
neuroimaging\cite{millman2007analysis},
geographic information systems\cite{shi2007python}, and
education\cite{backer2007computational, myers2007pythona}.

@@

In the early workshops, recurrent topics reflected the state of development, with
emphasis being placed on the underlying array package, plotting,
parallel processing, acceleration / wrapping, and user interfaces.  By
2004, a significant shift occurred towards application of SciPy to
scientific problems.  The event also started to draw in more keynote
speakers from outside the community, such as Guido van Rossum (creator
of Python, 2006), Ivan Krstić (One Laptop per Child, 2007), Alex
Martelli (Google, 2008), and Peter Norvig (Google Research, 2009).
The SciPy conference went from being a small gathering of
core developers, to a multi-location event, with increased funding, a
published proceedings, and scholarships for attending students (some
of whom are authors on this paper).  In 2010, the conference grew even
further, with an additional event added in India, and the US
conference—now with parallel tracks—moving to Austin, Texas.
While the US SciPy conference kept growing (by ... it had multiple
tracks, student sponsorship, proceedings, and funding), satellite
events were being organized by volunteers elsewhere, such as EuroSciPy
(2008–) and SciPy India (2009–).
Special sessions and minisymposia dedicated to scientific Python began
appearing at many events.
For example, a three-part minisymposium organized for CSE 2009 was
featured in SIAM News\cite{siamcse09}.

When Paul Dubois edited the May/June 2007 special issue of the 
IEEE publication Computing in Science and Engineering (CiSE), 
Python already had a strong enough presence in science and engineering
that the editors solicited the endeavor. However, Python was still sufficiently
niche that the average reader
would need additional information in order to decide whether it would
``be useful in their own work.''
The follow-up CiSE March/April 2011 Python for Scientists and Engineers
special issue\cite{millman2011python}, focused more on the core parts
of the scientific Python ecosystem\cite{perez2011python} including
NumPy\cite{vanderwalt2011numpy}, Cython\cite{behnel2011cython},
and Mayavi\cite{ramachandran2011mayavi}.

Python has become so pervasive, there are now domain-specific special issues.
For example, Frontiers in Neuroinformatics has a collection of 25 articles---covering
topics including modeling and simulation, data collection, electrophysiology, visualization,
as well as stimulus generation and presentation---called Python in
neuroscience\cite{python-FIN}.

In 2012, Perry Greenfield, John Hunter, Jarrod Millman, Travis Oliphant,
and Fernando Pérez founded NumFOCUS\cite{numfocus},
 a 501(c)3 public charity whose mission is to
``promote sustainable high-level programming languages, open code development,
and reproducible scientific research.''
While NumFOCUS is language agnostic, many of the early sponsored projects
came from the scientific Python stack.
Today it has over 25 sponsored projects including NumPy, SciPy, IPython, and
Matplotlib.
Among its other projects, NumFOCUS organizes a global network of 
community driven educational programs, meetups, and conferences
called PyData.
